Marker_identification_Benchmark

### 2024-05-23 Demo_marrow_facs.Rmd
240523 pipeline-development

# Install & Load Packages
```{r}
# Install necessary tools if not already present
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
# Define package lists
bioc_packages <- c("Seurat",
                   "clusterProfiler", "AnnotationDbi", "ReactomePA", "org.Mm.eg.db","gprofiler2")
github_packages <- c("satijalab/seurat-wrappers")
# Note: repeating packages from LMD_function.R as knitr does not install them properly
cran_packages <- c("igraph","ggplot2", "cowplot", "RColorBrewer",
                   "data.table", "dplyr", "patchwork",
                   "pheatmap", "ggplotify", "ggraph",
                   "ClusterR", "Rcpp", "RcppArmadillo", "tictoc")

# Install packages
sapply(cran_packages, function(pkg) if(!requireNamespace(pkg, quietly = TRUE)){install.packages(pkg)})
sapply(bioc_packages, function(pkg) if (!requireNamespace(pkg, quietly = TRUE)) BiocManager::install(pkg))
sapply(github_packages, function(pkg) if (!requireNamespace(strsplit(pkg, "/")[[1]][2], quietly = TRUE)) remotes::install_github(pkg))
library("Seurat")
library("SeuratWrappers")
library("pheatmap")
library("ggplot2")
library("ggplotify")
```
# Create a folder to save results
```{r}
folder_path = "marrow_result"
if (!dir.exists(path)) {
  # If it doesn't exist, create it
  dir.create(path)
} else {
}
```

# Preparing input data
```{r}
# tiss <- readRDS("/data/ruiqi/local_marker/dataset/tabular_muris/marrow_facs.rds")
```

@@

# Running LMD in one step

```{r}
tic()
lmd_result = LMD(dat, feature_space)
toc()
```

@@

# Hierarchical Tree
clustering_method = "average"
gene_hree = hclust(dist, method = clustering_method)

# Obtain Gene modules
gene_partition = dynamicTreeCut::cutreeDynamic(
  dendro = gene_hree, 
  distM = as.matrix(dist), deepSplit = 2,
  pamStage = TRUE, pamRespectsDendro = TRUE,
  minClusterSize = 5)
names(gene_partition) = labels(dist)

# Re-name gene modules based on the order of htree
gene_partition = as.factor(gene_partition)
module_order = order(sapply(levels(gene_partition), function(mod) {
  median(which(gene_partition[gene_hree$order] == mod))
}))
levels(gene_partition)[module_order] = seq(nlevels(gene_partition))
gene_partition = factor(gene_partition,levels = seq(nlevels(gene_partition)))
rm(module_order)
```

### Visualize Heatmap
```{r}
p = pheatmap(as.matrix(dist),
              annotation_col = data.frame(Module = gene_partition), 
              cluster_rows = gene_hree, 
              cluster_cols = gene_hree,
              ann_colors = list(Module = setNames(colorRampPalette(brewer.pal(12, "Paired"))(nlevels(gene_partition)),levels(gene_partition))),
              treeheight_row = 0,
              col=colorRampPalette(c("firebrick3", "white"))(99),
              show_colnames = FALSE,
              show_rownames = FALSE,
              main = paste(distance_method,
                           clustering_method,sep = "-"),
              silent = TRUE )
as.ggplot(p)
```

## Step2: Computing Per-Cell Module Activity Scores
```{r}
knn = 10; major_vote = 5
A = Symmetric_KNN_graph(knn = knn, feature_space = feature_space)$'adj_matrix'
cell_block_loop = lapply(1:100,function(loop){
  GMM_subsampling(seed = loop, gene_partition, expr_dat = dat, cell_kNN_graph = A, major_vote)
})
cell_block_prop = Reduce(`+`, cell_block_loop) / length(cell_block_loop)
cell_block = cell_block_prop
```

## Step3: Identify Celltype-specific activated Modules
```{r}
cell_label = as.factor(tiss$cell_ontology_class)
abbrev = levels(cell_label)
abbrev[grepl("T cell",abbrev)] = "T cell"
abbrev[grepl("natural killer",abbrev)] = "NK cell"
abbrev[abbrev == "B cell"] = "Cd3e+ Klrb1+ B cell"
abbrev[grepl("Slamf1-positive",abbrev)] = "Slamf1+ MPC"
abbrev[grepl("Slamf1-negative",abbrev)] = "Slamf1- MPC"
abbrev[grepl("common lymphoid",abbrev)] = "CLP"
abbrev[grepl("megakaryocyte",abbrev)] = "MEP"
abbrev[grepl("hematopoietic",abbrev)] = "HSC"
abbrev[grepl("hematopoietic",abbrev)] = "HSC"
abbrev[grepl("granulocyte monocyte",abbrev)] = "GMP"
names(abbrev) = levels(cell_label)
levels(cell_label) = abbrev[levels(cell_label)]

# Create one-hot vector of Cell Type
one_hot_matrix <- sapply(unique(cell_label), function(x) as.integer(cell_label == x))
colnames(one_hot_matrix) = unique(cell_label)
rownames(one_hot_matrix) = names(cell_label)

# Calculate Jaccard index between module activity & cell type
jaccard_index <- apply(cell_block, 2, function(module_col) {
  apply(one_hot_matrix, 2, function(one_hot_col) {
    1 - philentropy::distance(rbind(module_col,one_hot_col), method = "jaccard", mute.message=TRUE)
  })
})
```

### Annotate Each Celltype-specific activated Modules with its corresponding Celltypes
```{r}
module_celltype = apply(jaccard_index,2,function(x){paste(names(which(x > 0.4)),collapse = ",")})
```

## Transferability of Modules
### Load Data & Calculate Module Activity
```{r}
if (!file.exists('droplet_Marrow_seurat_tiss.Robj')) {
   # If load url doesn't work, please download it and load from your working directory
  options(timeout=6000)
  download.file("https://figshare.com/ndownloader/files/13092380",
                destfile = 'droplet_Marrow_seurat_tiss.Robj', method = 'libcurl')
 }
load('droplet_Marrow_seurat_tiss.Robj')
tiss_droplet <- UpdateSeuratObject(tiss)
saveRDS(tiss_droplet, file = "processed_droplet_Marrow_srat.rds")

feature_space_droplet = as.matrix(tiss_droplet@reductions$pca@cell.embeddings[,1:n_dim])
dat_droplet = as.matrix(tiss_droplet[["RNA"]]@data)

knn = 10; major_vote = 5
A = Symmetric_KNN_graph(knn = knn, feature_space = feature_space_droplet)$'adj_matrix'
cell_block_loop = lapply(1:100,function(loop){
  GMM_subsampling(seed = loop, gene_partition, expr_dat = dat_droplet, cell_kNN_graph = A, major_vote)
})
cell_block_prop = Reduce(`+`, cell_block_loop) / length(cell_block_loop)
cell_block_droplet = cell_block_prop
```

### Calculate Jaccard Index
```{r}
cell_label = tiss_droplet$cell_ontology_class
cell_label[!cell_label %in% names(abbrev)] = NA
cell_label = setNames(abbrev[cell_label],names(cell_label))

one_hot_matrix_droplet <- sapply(unique(na.omit(cell_label)), function(x) as.integer(cell_label == x))
one_hot_matrix_droplet[is.na(one_hot_matrix_droplet)] = 0

jaccard_index_droplet <- apply(cell_block_droplet, 2, function(module_col) {
  apply(one_hot_matrix_droplet, 2, function(one_hot_col) {
    1 - philentropy::distance(rbind(module_col,one_hot_col), method = "jaccard", mute.message=TRUE)
  })
})
```
## Save Intermediate data
```{r}
save(gene_partition,dist,gene_hree,cell_block,cell_block_droplet,one_hot_matrix,one_hot_matrix_droplet,jaccard_index,jaccard_index_droplet,file = file.path(folder_path, "module_result.rdata"))

saveRDS(tiss, file = "processed_facs_Marrow_srat.rds")
```

## Step4: Module Enrichment Analysis
### GO enrichment Analysis
```{r}
library("clusterProfiler")
library("ReactomePA")
library("org.Mm.eg.db")
library("AnnotationDbi")

# Set all genes(after filtering) in the expression matrix as background
universe_df = data.frame("symbol" = rownames(dat),"entrez" = mapIds(org.Mm.eg.db, keys=rownames(dat), column="ENTREZID", keytype="SYMBOL"))

# Filter GO terms based on p values < 0.05
ego_result <- lapply(1:nlevels(gene_partition), function(i){
  enrichGO(gene = names(gene_partition)[gene_partition == i],
           OrgDb = 'org.Mm.eg.db', # mouse
           keyType = "SYMBOL",
           ont = "ALL",
           pAdjustMethod = "BH",
           universe = universe_df$symbol,
           pvalueCutoff = 0.05)
})

# Only keep modules which has at lease one GO term having adjusted p-value < 0.05
top_go_clusters <- which(sapply(ego_result, function(x)
  (x@result$p.adjust[1] < 0.05)))

# Annotate each modules with top5 GO terms
top_enrich_go <- lapply(top_go_clusters, function(i) ego_result[[i]]@result[1:5,])
```

### Reactome enrichment Analysis

```{r}
library("ReactomePA")
library("org.Mm.eg.db")

# Set all genes(after filtering) in the expression matrix as background
universe_df = data.frame("symbol" = rownames(dat),"entrez" = mapIds(org.Mm.eg.db, keys=rownames(dat), column="ENTREZID", keytype="SYMBOL"))
universe_df = universe_df[!is.na(universe_df$entrez),]

# Filter Pathway based on p values < 0.05
epathway_result <- lapply(1:nlevels(gene_partition), function(i){
  enrichPathway(gene=universe_df[universe_df$symbol %in% names(gene_partition)[gene_partition == i],"entrez"],
                organism = "mouse",
                pAdjustMethod = "BH",
                pvalueCutoff = 0.05,
                universe = universe_df$entrez)
})

# Only keep modules which has at lease one Pathway having adjusted p-value < 0.05
top_pathway_clusters <- which(sapply(epathway_result, function(x) 
  (x@result$p.adjust[1] < 0.05)))

# Annotate each modules with top5 Pathway terms
top_enrich_pathway <- lapply(top_pathway_clusters, function(i) epathway_result[[i]]@result[1:5,])
```

## Step5: Generate Gene Module Annotation Table
```{r}
df_module = data.frame(Module = levels(gene_partition))
df_module[,"Gene"] = unlist(lapply(levels(gene_partition),function(x){paste(names(gene_partition)[gene_partition == x],collapse = ",")}))
df_module[,"celltype"] = module_celltype[levels(gene_partition)]
df_module[top_go_clusters,"GO_description"] = unlist(lapply(top_enrich_go,function(x){
  des = x[["Description"]]
  paste(des[!is.na(des)],collapse = ",")
}
))
df_module[top_pathway_clusters,"Pathway_description"] = unlist(lapply(top_enrich_pathway,function(x){
  des = x[["Description"]]
  paste(des[!is.na(des)],collapse = ",")
}))
df_module[is.na(df_module)] = ""
write.csv(df_module,file = file.path(folder_path, "module_description.csv"),row.names = FALSE)
```

## Cell Cycle Modules
### Reactome analysis
```{r}
library("ReactomePA")
library("org.Mm.eg.db")
universe_df = data.frame("symbol" = rownames(dat),"entrez" = mapIds(org.Mm.eg.db, keys=rownames(dat), column="ENTREZID", keytype="SYMBOL"))
universe_df = universe_df[!is.na(universe_df$entrez),]
module_cc = c(16,18,17,31,32)
epathway_result_cc <- lapply(module_cc, function(i){
  enrichPathway(gene=universe_df[universe_df$symbol %in% names(gene_partition)[gene_partition == i],"entrez"],
                organism = "mouse",
                pAdjustMethod = "BH",
                pvalueCutoff = 0.05,
                universe = universe_df$entrez)
})
names(epathway_result_cc) = module_cc
# Annotate each modules with top5 Pathway terms
top_enrich_pathway_cc <- do.call(rbind,lapply(1:length(epathway_result_cc), function(i) data.frame(epathway_result_cc[[i]]@result[1:5,],module = module_cc[i])))
top_enrich_pathway_cc <- top_enrich_pathway_cc[complete.cases(top_enrich_pathway_cc), ]
top_enrich_pathway_cc = top_enrich_pathway_cc %>% filter(p.adjust < 0.05)

write.csv(top_enrich_pathway_cc,file = file.path(folder_path, "reactome_pathway_cc.csv"))
```

### Annotate Cell Cycle
```{r}
mmus_s = gprofiler2::gorth(cc.genes.updated.2019$s.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name
mmus_g2m = gprofiler2::gorth(cc.genes.updated.2019$g2m.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name
tiss = CellCycleScoring(tiss, s.features = mmus_s, 
                        g2m.features = mmus_g2m, set.ident = FALSE)
cell_cycle_table = data.frame(tiss@meta.data[,c("S.Score","G2M.Score","Phase")],"Dataset" = "FACS")
```

### Transferablity
```{r}
tiss_droplet = CellCycleScoring(tiss_droplet, s.features = mmus_s, 
                        g2m.features = mmus_g2m, set.ident = FALSE)
cell_cycle_table = rbind(cell_cycle_table,data.frame(tiss_droplet@meta.data[,c("S.Score","G2M.Score","Phase")],"Dataset" = "Droplet"))

write.csv(cell_cycle_table,file = file.path(folder_path, "annotated_cell_cycle.csv"))
```

>>>>

# Install & Load Packages
```{r}
source("LMD_function.R")
# Define package lists
bioc_packages <- c("clusterProfiler", "AnnotationDbi", "ReactomePA", "org.Mm.eg.db","gprofiler2")
github_packages <- c("satijalab/seurat-wrappers");
names(github_packages) = strsplit(github_packages, "/")[[1]][2]
names(github_packages)[grepl("seurat-wrappers",names(github_packages))] = "SeuratWrappers"
# Install packages
sapply(bioc_packages, function(pkg) if (!requireNamespace(pkg, quietly = TRUE)) BiocManager::install(pkg))
sapply(1:length(github_packages), function(i) if (!requireNamespace(names(github_packages)[i], quietly = TRUE)) remotes::install_github(github_packages[i]))
lapply(c(bioc_packages,names(github_packages)), require, character.only = TRUE)
```

# Create a folder to save results
```{r}
folder_path = "marrow_result"
if (!dir.exists(path)) {
  # If it doesn't exist, create it
  dir.create(path)
} else {
}
```

# Preparing input data
```{r}
# tiss <- readRDS("/banach1/ruiqi/local_marker/LMD_data/tabular_muris/marrow_facs.rds")
```

@@

# Running LMD in one step
```{r}
tic()
lmd_result = LMD(dat, feature_space)
toc()
# lmd_result = LMD(dat, feature_space, highres = TRUE, max_time = 2^7)
```

@@

# Obtain Gene modules
gene_partition = Obtain_gene_partition(dist, clustering_method = "average", min_gene = 5)
```

### Visualize Heatmap
```{r}
p = Visualize_gene_heatmap(dist)
```

## Step2: Computing Per-Cell Module Activity Scores
```{r}
tiss = AddModuleActivityScore(tiss, gene_partition = gene_partition, do_local_smooth = TRUE)
```

### 2024-05-23 miscellaneous/miscellaneous_code/Marker_identfication_benchmark_tutorial.Rmd
240523 pipeline-development

# Load Data & Process
## Load Data
```{r}
# tiss <- readRDS("/data/ruiqi/local_marker/dataset/tabular_muris/marrow_facs.rds")
# DefaultAssay(tiss) <- "RNA"
```

```{r}
tissue_fullname = "Bone marrow"

@@

## Process
```{r}
DefaultAssay(tiss) <- "RNA"
n_dim = pmin(dim(tiss@reductions$pca@cell.embeddings)[2],30)
feature_space = as.matrix(tiss@reductions$pca@cell.embeddings[,1:n_dim])
dat = as.matrix(tiss[[DefaultAssay(tiss)]]@data)

@@

# Load data
data = pd.DataFrame(r.dat.T)  # counts
PC20 = pd.DataFrame(r.feature_space)
# Reference Cell
g = 8.6e-4  
S = pairwise_kernels(PC20, metric="rbf", gamma=g) 
median = np.median(S, axis=0)
start = int(np.argmin(median))
num_refcell = int(data.shape[1]*0.005)
dd_rcells = from_knn_dist(X=PC20, 
                          n_ret=num_refcell,
                          start=start, 
                          metric="rbf",
                          metric_params={"gamma": g})
S = pairwise_similarities(PC20,
                          query=dd_rcells,
                          metric="rbf",
                          metric_params={"gamma": g})
dd_rcells = data.index[dd_rcells]
escores = calculate_escores(data.values, query=dd_rcells, S=S)

# semi_ranks = escores.abs().rank(ascending=False, method="dense").min(1)

P = permute(data.values)
pscores = calculate_escores(P, query=dd_rcells, S=S)
pvals = pvals_per_cell(escores,pscores,ret = "q")
# semi_ranks = pvals.min(1).rank(ascending=True, method="dense")

@@

  df = roc_df %>% select(c("Method","AUROC")) %>% distinct()
  df$'gt_set' = names(gt_list)[i]
  df
}))
df$Method = as.factor(df$Method)
df$gt_set = factor(df$gt_set,levels = names(gt_list))
write.table(df,file = file.path(folder_path, paste0(tissue_name,"_auroc.csv")),row.names = FALSE)
```

```{r}
# Concordance table
do.call(cbind,lapply(gt_list[c(2,21)],function(true_marker){
  df_benchmark$"gt" = 0
  df_benchmark[true_marker,"gt"] = 1
  apply(df_benchmark,2,function(x){
    sum(df_benchmark[order(x),"gt"][1:100])
  })
}))

@@

# Prop of recovered cell types
recovery_rate =data.frame(sapply(c(colnames(df_benchmark)),function(method){
  top_gs = rownames(df_benchmark)[order(df_benchmark[,method])]
  top_n = 1:300
  scores = sapply(top_n, function(top){
    features = top_gs[1:top]
    mean(colSums(recovery_mtx[features,,drop = FALSE]) > 0)
  }) 
}) )
recovery_rate$'TopGene' = 1:nrow(recovery_rate)
```

# Sensitivity Test (Try different K)
```{r}
# RunLMD under differen K
source("./LMD_function.R", echo = F)
score_df = do.call(cbind,lapply(c(3:10,20,30,40,50),function(kNN){
  try({
    res = LMD(dat,feature_space,knn = kNN)
    return(res$cumulative_score)
  },silent = TRUE)
}))
colnames(score_df) = paste0("kNN",c(3:10,20,30,40,50))
write.table(score_df,file = file.path("marrow_result", paste0(tissue_name,"_different_knn.csv")),row.names = TRUE)
```
```{r}
# Obtain AUROC
score_ranked <- data.frame(score_df) %>% mutate(across(everything(), rank))

>>>>

# Load Data & Process
## Load Data
```{r}
tissue_fullname = "Bone marrow"

@@

## Process
```{r}
# tiss <- readRDS("/banach1/ruiqi/local_marker/dataset/tabular_muris/marrow_facs.rds")
# DefaultAssay(tiss) <- "RNA"
# tiss = tiss %>% NormalizeData(1e6) %>%
#   ScaleData(verbose = FALSE) %>% FindVariableFeatures() %>% RunPCA(npcs = 50, verbose = FALSE)
```

```{r}
DefaultAssay(tiss) <- "RNA"
n_dim = 20
feature_space = as.matrix(tiss@reductions$pca@cell.embeddings[,1:n_dim])
dat = as.matrix(tiss[[DefaultAssay(tiss)]]@data)

@@

# Load data
data = pd.DataFrame(r.dat.T)  # counts
PC_embed = pd.DataFrame(r.feature_space)
# Reference Cell
g = 8.6e-4  
S = pairwise_kernels(PC_embed, metric="rbf", gamma=g) 
median = np.median(S, axis=0)
start = int(np.argmin(median))
num_refcell = int(data.shape[1]*0.005)
dd_rcells = from_knn_dist(X=PC_embed, 
                          n_ret=num_refcell,
                          start=start, 
                          metric="rbf",
                          metric_params={"gamma": g})
S = pairwise_similarities(PC_embed,
                          query=dd_rcells,
                          metric="rbf",
                          metric_params={"gamma": g})
dd_rcells = data.index[dd_rcells]
escores = calculate_escores(data.values, query=dd_rcells, S=S)
# semi_ranks = escores.abs().rank(ascending=False, method="dense").min(1)
P = permute(data.values)
pscores = calculate_escores(P, query=dd_rcells, S=S)
pvals = pvals_per_cell(escores,pscores,ret = "q")
# semi_ranks = pvals.min(1).rank(ascending=True, method="dense")

@@

  df = roc_df %>% select(c("Method","AUROC")) %>% distinct()
  df$'gt_set' = names(gt_list)[i]
  df
}))
df$Method = as.factor(df$Method)
df$gt_set = factor(df$gt_set,levels = names(gt_list))
write.table(df,file = file.path(folder_path, paste0(tissue_name,"_auroc.csv")),row.names = FALSE)
```
# Concordance table
```{r}
do.call(cbind,lapply(gt_list[c(2,21)],function(true_marker){
  df_benchmark$"gt" = 0
  df_benchmark[true_marker,"gt"] = 1
  apply(df_benchmark,2,function(x){
    sum(df_benchmark[order(x),"gt"][1:100])
  })
}))

@@

# Prop of recovered cell types
recovery_rate =data.frame(sapply(c(colnames(df_benchmark)),function(method){
  top_gs = rownames(df_benchmark)[order(df_benchmark[,method])]
  top_n = 1:300
  scores = sapply(top_n, function(top){
    features = top_gs[1:top]
    mean(colSums(recovery_mtx[features,,drop = FALSE]) > 0)
  }) 
}) )
recovery_rate$'TopGene' = 1:nrow(recovery_rate)
```
## Plot
```{r}
recovery_rate <- reshape2::melt(recovery_rate,id.vars = "TopGene",value.name = "prop",variable.name = "Method")
recovery_rate = recovery_rate %>% filter(Method!="wilcox_no_filter")
recovery_rate$Method = ifelse(recovery_rate$Method %in% names(method_fullname),method_fullname[recovery_rate$Method],recovery_rate$Method)
recovery_rate$'color' = coldef[recovery_rate$Method]

ggplot(recovery_rate, aes(x = TopGene, y = prop, color = Method, group = Method)) + 
  geom_line() +
  scale_color_manual(values = score_measure %>% filter(Method!="All genes") %>% select(c("Method","color")) %>% distinct() %>% tibble::deframe()) + ylab("Prop. of recovered cell types")
```

# Sensitivity Test
## Different kNN
```{r}
# RunLMD under differen K
source("./LMD_function.R", echo = F)
score_df = do.call(cbind,lapply(c(3:10,20,30,40,50),function(kNN){
  try({
    res = LMD(dat,feature_space,knn = kNN)
    return(res$cumulative_score)
  },silent = TRUE)
}))
colnames(score_df) = paste0("kNN",c(3:10,20,30,40,50))
write.table(score_df,file = file.path("marrow_result", paste0(tissue_name,"_different_knn.csv")),row.names = TRUE)
```
```{r}
# Obtain AUROC
score_ranked <- data.frame(score_df) %>% mutate(across(everything(), rank))
